---
title: "Food consumption and quality upgrading in Tanzania"
subtitle: "First-stage analysis — automated report"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: kable
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(knitr); library(dplyr); library(readr)
tab <- function(f) readr::read_csv(file.path("output/tables", f), show_col_types = FALSE)
fig <- function(f) knitr::include_graphics(file.path("output/figures", f))
num <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")
```

> **This report is generated from the outputs of `01_first_stage_analysis.R`.**
> Run that script first (it writes `output/tables/*.csv` and `output/figures/*.png`),
> then knit this file. All tables/figures below are read from those outputs, so
> the report always matches the latest run.

# Objective and framework

We study animal-source-food (ASF) consumption and **quality upgrading** in
Tanzania using the National Panel Survey (NPS), waves 3–5 pooled. For each item
the budget identity is exact:

$$\ln(\text{expenditure}) = \ln(\text{quantity}) + \ln(\text{unit value})$$

Regressing each term on log real expenditure per adult equivalent ($W$) gives the
Deaton (1988) decomposition $\;\varepsilon_{exp} = \varepsilon_{qty} + \varepsilon_{quality}$,
and we test $H_0:\varepsilon_{quality}=0$ vs $H_1:\varepsilon_{quality}>0$.
The unit value (price paid per kg/litre/piece) is the proxy for quality.

**Items:** goat meat, beef, pork, chicken & poultry, eggs, fresh milk
(*processed milk is not in the master files*). Welfare = real total expenditure
per adult equivalent. Descriptives are survey-weighted; regressions are
unweighted with **region-clustered** standard errors and **wave fixed effects**.

# Sample and households

```{r}
kable(tab("01_household_summary.csv"), digits = c(0,0,2,2,1,1,1,0),
      caption = "Household summary by wave (weighted)")
```

```{r}
fig("fig6_household_composition.png")
```

# Descriptive statistics by item

Means over consumers, at **adult-equivalent** level; unit value is the median.

```{r}
kable(tab("02_item_summary.csv"), digits = c(0,1,2,0,2,0,0),
      caption = "Item summary (household and per-adult-equivalent)")
```

```{r, fig.show='hold', out.width="49%"}
fig("fig7_participation.png"); fig("fig2_unitvalue_quintiles.png")
```

## Where consumed quantity comes from (purchased / own / gifts)

```{r}
kable(tab("02c_quantity_source.csv"), digits = 2,
      caption = "Mean quantity per AE by source")
```

```{r, fig.show='hold', out.width="49%"}
fig("fig16_quantity_source.png"); fig("fig17_source_by_livestock.png")
```

Beef and pork are almost entirely **purchased**; **chicken is largely received
as gifts**; goat meat, eggs and milk have a sizeable **own-production** share.
Livestock owners self-provision (own production), non-owners buy or receive gifts.

## By livestock ownership

```{r}
kable(tab("02b_item_by_livestock.csv"), digits = c(0,0,1,2,0,0),
      caption = "Item consumption by livestock ownership")
fig("fig18_expenditure_by_livestock.png")
```

# Consumption gradients across the welfare distribution

Top/bottom ratios (richest vs poorest group), per adult equivalent. Both the
**quintile (Q5/Q1)** and the **quartile (Qt4/Qt1)** versions are reported; cells
with fewer than 30 consumers are blanked. For rural/urban the welfare groups are
defined **within area**.

```{r}
kable(tab("04_top_bottom_ratios.csv"), digits = c(0,0,0,0,2,2,2),
      caption = "Top/bottom ratios — quantity, expenditure, unit value")
```

```{r, fig.show='hold', out.width="49%"}
fig("fig1_Q5Q1_ratios.png"); fig("fig10_Q5Q1_by_area.png")
```

```{r, fig.show='hold', out.width="49%"}
fig("fig8_quantity_quintiles.png"); fig("fig9_expenditure_quintiles.png")
```

```{r}
fig("fig20_unitvalue_rural_urban.png")
```

As income rises households buy **more** and pay a **modest premium per unit**
(quality), the latter strongest for chicken.

# Elasticity decomposition (main result)

```{r}
kable(tab("05_elasticities.csv") %>%
        select(item_label, n, eps_expenditure, eps_quantity, eps_quality,
               quality_p_1side, reject_H0_5pct),
      digits = c(0,0,3,3,3,4,0),
      caption = "Expenditure elasticity = quantity + quality (region-clustered SE)")
```

```{r, fig.show='hold', out.width="49%"}
fig("fig4_elasticity_decomposition.png"); fig("fig15_quality_forest.png")
```

**H₀ is rejected for every item** — quality upgrading is present. But the income
response runs **mainly through quantity** (~80–88%); the quality component, while
significant, is smaller. Expenditure elasticities (0.37–0.55) are
*intensive-margin, conditional-on-purchase* elasticities, hence below textbook
unconditional ASF values.

## By subgroup: rural / urban / livestock

```{r}
kable(tab("05b_elasticities_by_subgroup.csv"),
      digits = c(0,0,3,3,3,3,4),
      caption = "Elasticities by subgroup (pooled, item FE)")
```

# Does quality matter more as income rises?

```{r}
kable(tab("06_quality_by_quintile.csv"), digits = c(0,3,3,2,4,0),
      caption = "Quality elasticity by welfare quintile")
fig("fig5_quality_by_quintile.png")
```

The quality elasticity is **positive and significant in every quintile**,
including the poorest, and is roughly flat — so there is **no clean income
threshold** at which quality "switches on".

## The quantity → quality shift (indicative)

Quality's *share* of the expenditure response by welfare group, and the income
(welfare per AE) at which that share stops falling and starts rising.

```{r}
kable(tab("08_quality_share_by_group.csv"), digits = c(0,0,3,3,1),
      caption = "Quality share of the expenditure elasticity")
kable(tab("08b_shift_income_by_wave.csv"), digits = c(0,0,0,0),
      caption = "Indicative income where the shift begins (welfare per AE, by wave)")
fig("fig19_quality_share.png")
```

The profile is **shallow**: quantity dominates throughout; quality's share is
lowest in the lower-middle of the distribution and rises towards the top. Read
the "shift income" as indicative, not a sharp turning point.

# Robustness

```{r}
kable(tab("09_elasticities_by_wave.csv"), digits = c(0,0,3,3,3),
      caption = "Elasticities estimated separately by wave")
```

Additional checks (in the script / `FINAL_REPORT.md`): the adding-up identity
holds exactly; the pooled quality elasticity is stable to outlier trimming
(0.071–0.089); HC1 and region-clustered SEs agree; per-item estimates are robust
while the pooled estimate requires item fixed effects.

```{r, fig.show='hold', out.width="49%"}
fig("fig12_engel_curves.png"); fig("fig13_quality_gradient.png")
```

# Caveats

- Elasticities are **intensive-margin, conditional on purchase**, w.r.t. a
  welfare proxy (total expenditure per AE) — not structural income elasticities.
- **Unit value = quality** is an assumption; it also reflects bulk discounts,
  spatial price differences and measurement error (mitigated by region FE and
  trimming).
- Cross-wave welfare **levels** are not comparable (deflator base differs) →
  within-wave groups + wave FE; absolute monetary thresholds are only indicative.
- Thin items (pork, goat) and the urban subsample have small cells; ratios below
  30 consumers are suppressed.

# Conclusions

1. ASF are normal goods on the intensive margin; income raises consumption
   **mainly via quantity**.
2. **Quality upgrading is statistically robust** for all six items (H₀ rejected)
   but **economically modest** (~12–25% of the response) and **present across the
   whole distribution**, not only at the top.
3. Strongest quality response: **chicken & poultry**; weakest: **beef**.
   Quality upgrading is somewhat stronger in **rural** areas.
